---
title: "NYC Shootings incidents and Criminality <br> Data preprocessing"
author: "Sebastiano Quintavale - 878500"
output:
    html_document:
        toc: true
        toc_float: true
        theme: cerulean
        highlight: tango
---


## 1. Introduction

This document describes the data preprocessing phase of the project, which involves the cleaning and transformation of the original data into a format suitable for analysis.

The two datasets used for the analysis are as follows:

- **NYPD Shooting Incident Data** - This dataset contains information about shooting incidents in New York City from 2006 to 2023. It is available on the NYC Data Portal at the following [resource](https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8).
- **NYPD Arrest Data** - This dataset provides information about arrests in New York City from 2006 to 2023. It is available on the NYC Data Portal at the following [resource](https://data.cityofnewyork.us/Public-Safety/NYPD-Arrest-Data-Historic-/8h9b-rp9u).

The goal of the preprocessing is to merge these two data sources and aggregate the shooting and arrest counts by month. Each entry in the resulting dataset will contain the number of **shootings** and **arrests** for different **crime types**, categorized by year, month, and New York City borough.

```{r setup, echo=FALSE, include=FALSE}

# Don't display code in the final document
knitr::opts_chunk$set(echo=FALSE)

# Load R script sources
source("src/init.R")

```

```{r names}

# Columns to manipulate

SHOOTINGS_COLS <- c("INCIDENT_KEY", "OCCUR_DATE",  "BORO")
ARRESTS_COLS   <- c("ARREST_KEY",   "ARREST_DATE", "ARREST_BORO", "PD_DESC")

# Name mappings

BOROUGH_LETTER_MAP <- c(
    "B" = "BRONX",
    "K" = "BROOKLYN",
    "M" = "MANHATTAN",
    "Q" = "QUEENS",
    "S" = "STATEN ISLAND"
)


BOROUGH_LOWERCASE_MAP <- c(
    "BRONX"         = "Bronx",
    "BROOKLYN"      = "Brooklyn",
    "MANHATTAN"     = "Manhattan",
    "QUEENS"        = "Queens",
    "STATEN ISLAND" = "StatenIsland"
)

CRIME_TYPES <- list(

    # Aggressione
    Assault = c(
        "ASSAULT 3",                 "ASSAULT 2,1,UNCLASSIFIED", 
        "ASSAULT 2,1,PEACE OFFICER", "ASSAULT POLICE/PEACE OFFICER"
    ),

    # Furto con scasso
    Burglary = c(  
        "BURGLARY,UNCLASSIFIED,UNKNOWN TIME", "BURGLARY,UNCLASSIFIED,UNKNOWN",
        "BURGLARY,RESIDENCE,NIGHT",           "BURGLARY,COMMERCIAL,DAY"
    ),

    # Armamentario di droga
    DrugParaphernalia = c(  
        "DRUG PARAPHERNALIA,   POSSESSES OR SELLS 2", "DRUG PARAPHERNALIA,   POSSESSE",
        "DRUG PARAPHERNALIA,   POSSESSES OR SELLS 1"
    ),

    # Furto di proprietà
    Larceny= c(
        "LARCENY,PETIT FROM OPEN AREAS,UNCLASSIFIED", "LARCENY,PETIT FROM OPEN AREAS,",           "LARCENY,GRAND FROM OPEN AREAS,UNCLASSIFIED",
        "LARCENY,GRAND FROM OPEN AREAS, UNATTENDED",  "LARCENY,GRAND OF AUTO",                    "LARCENY,GRAND FROM PERSON,UNCLASSIFIED",
        "LARCENY,GRAND FROM PERSON,UNCL",             "LARCENY,GRAND FROM OPEN AREAS,",           "LARCENY,GRAND BY EXTORTION",
        "AGGRAVATED GRAND LARCENY OF ATM",            "LARCENY,GRAND BY CREDIT CARD USE",         "LARCENY,PETIT BY ACQUIRING LOS",
        "LARCENY,GRAND BY ACQUIRING LOS",             "LARCENY,GRAND FROM BUILDING,UNCLASSIFIED", "LARCENY,GRAND BY THEFT OF CREDIT CARD",
        "LARCENY,GRAND FROM BUILDING (NON-RESIDENCE) UNATTENDED"
    ),

    # Violation of firearm license
    LicensingFirearm = c( 
        'CRIMINAL DISPOSAL FIREARM 1 & 2', "CRIMINAL DISPOSAL FIREARM 1 &", "CRIMINAL DISPOSAL FIREARM 1",
        "LICENSING FIREARMS",              "FIREARMS LICENSING LAWS"
    ),


    # Possesso di marijuana
    MarijuanaPossession = c(  
        "MARIJUANA, POSSESSION 4 & 5", "MARIJUANA, SALE 4 & 5",    "MARIJUANA, POSSESSION 1, 2 & 3",
        "MARIJUANA, POSSESSION",       "MARIJUANA, SALE 1, 2 & 3", "UNLAWFUL SALE SYNTHETIC MARIJUANA"
    ),

    # Danneggiamento
    Mischief = c( 
        "MISCHIEF,CRIMINAL     UNCLASSIFIED 4TH DEG", "MISCHIEF,CRIMINAL,    UNCL 2ND",             "MISCHIEF,CRIMINAL,    UNCL 2ND DEG 3RD DEG",
        "CRIMINAL MISCHIEF,UNCLASSIFIED 4",           "CRIMINAL MISCHIEF 4TH, GRAFFITI",            "CRIMINAL MISCHIEF 4TH, GRAFFIT",
        "MISCHIEF, CRIMINAL 4, OF MOTOR VEHICLE",     "MISCHIEF, CRIMINAL 3 & 2, OF MOTOR VEHICLE", "MISCHIEF 1,CRIMINAL,EXPLOSIVE",
        "MISCHIEF, CRIMINAL 4, BY FIRE",              "MISCHIEF, CRIMINAL 3 & 2, OF M",             "MISCHIEF, CRIMINAL 4, OF MOTOR",
        "MISCHIEF,CRIMINAL     UNCLASSI"
    ),

    # Omicidio
    Murder = c("MURDER,UNCLASSIFIED"),

    # Rapina
    Robbery = c( 
        "ROBBERY,UNCLASSIFIED,OPEN AREAS",           "ROBBERY,OPEN AREA UNCLASSIFIED", "ROBBERY,UNCLASSIFIED,OPEN AREA",
        "ROBBERY,CARJACKING OF MV OTHER THAN TRUCK", "ROBBERY,CAR JACKING",            "ROBBERY,GAS STATION"
    ),
    
    # Possesso di sostanze
    SubstancePossession = c(  
        "CONTROLLED SUBSTANCE, POSSESSION 7", "CONTROLLED SUBSTANCE,INTENT TO SELL 3",  "CONTROLLED SUBSTANCE,SALE 3",
        "CONTROLLED SUBSTANCE, POSSESSI",     "CONTROLLED SUBSTANCE, INTENT TO SELL 5", "CONTROLLED SUBSTANCE,INTENT TO",
        "CONTROLLED SUBSTANCE, POSSESSION 5", "CONTROLLED SUBSTANCE,POSSESS. 1",        "CONTROLLED SUBSTANCE, POSSESSION 4",
        "CONTROLLED SUBSTANCE,POSSESS. 2",    "CONTROLLED SUBSTANCE, SALE 5",           "CONTROLLED SUBSTANCE,POSSESS.",
        "CONTROLLED SUBSTANCE, INTENT T",     "CONTROLLED SUBSTANCE,SALE 1",            "CONTROLLED SUBSTANCE,POSSESS. 3",
        "CONTROLLED SUBSTANCE,SALE 2",        "CONTROLLED SUBSTANCE, SALE 4",           "CONTROLLED SUBSTANCE,POSSESS. OF PROCURSERS"
    ),

    # Furto di servizi
    Theft = c( 
        "THEFT OF SERVICES, UNCLASSIFIED",     "THEFT OF SERVICES, UNCLASSIFIE",  "THEFT,RELATED OFFENSES,UNCLASSIFIED",
        "THEFT OF SERVICES- CABLE TV SERVICE", "THEFT,RELATED OFFENSES,UNCLASS"
    ),

    

    # Possesso di armi
    WeaponPossession = c( 
        "WEAPONS, POSSESSION, ETC",   "WEAPONS POSSESSION 3",                    "WEAPONS POSSESSION 1 & 2",
        "WEAPONS,MFR,TRANSPORT,ETC.", "WEAPONS,PROHIBITED USE",                  "WEAPONS,DISPOSITION OF",
        "WEAPONS DISPOSITION OF",     "WEAPONS,PROHIBITED USE IMITATION PISTOL", "WEAPONS,PROHIBITED USE IMITATI"
    )

)


```

## 2. Shootings data

We start by loading  the **shootings dataset** [`shootings.csv`](/data/shooting.csv).  

```{r load shootings}

shootings <- read.csv(PATHS$Shootings)

head(shootings, 10)
```

We only keep the columns of interest, which are `INCIDENT_KEY`, `OCCUR_DATE`, and `BORO`, as we are primarily interested in identifying the time and location of the shootings to perform temporal and spatial aggregation. Although we could use the geocoordinates of the shootings for a custom spatial aggregation, their usage is not recommended in the [dataset footnotes](https://data.cityofnewyork.us/api/views/833y-fsy8/files/e4e3d86c-348f-4a16-a17f-19480c089429?download=true&filename=NYPD_Shootings_Incident_Level_Data_Footnotes.pdf). Therefore, a spatial aggregation by Borough is performed.

```{r shootings type casting}

# Keep columns of interest
shootings <- shootings[, SHOOTINGS_COLS]

# Cast to date and time type
shootings$OCCUR_DATE <- as.Date(shootings$OCCUR_DATE, format = "%m/%d/%Y")

# Cast Borough to factor
shootings$Borough <- as.factor(shootings$BORO)
shootings$BORO    <- NULL

# Extract time features
shootings$Year  <- as.numeric(format(shootings$OCCUR_DATE, "%Y"))
shootings$Month <- as.numeric(format(shootings$OCCUR_DATE, "%m"))

# Save the first and last date
date.first <- min(shootings$OCCUR_DATE)
date.last  <- max(shootings$OCCUR_DATE)

shootings$OCCUR_DATE <- NULL

head(shootings)
```

Here a summary of the shootings dataset.

```{r}
summary(shootings)
```

We can now aggregate the shootings data by year, month, and borough to obtain a count of shootings for each month and borough.

```{r shootings month aggregation}

# Create a new dataset counting how many arrests in each month
shootings_month <- aggregate(
    INCIDENT_KEY ~ Borough + Month + Year,
    data = shootings,
    FUN  = length
)

# Change count colname
names(shootings_month)[4] <- "Shootings"

# Create a dummy dataset to fill missing months with no shooting
shootings_month_dummy <- expand.grid(
    OCCUR_DATE     = seq(date.first, date.last, by = "month"),
    Borough        = levels(shootings$Borough),
    ShootingsDummy = 0
)

# Cast to the appropriate type
shootings_month_dummy$OCCUR_DATE <- as.Date(shootings_month_dummy$OCCUR_DATE, format="%m/%d/%Y")
shootings_month_dummy$Year       <- as.numeric(format(shootings_month_dummy$OCCUR_DATE, "%Y"))
shootings_month_dummy$Month      <- as.numeric(format(shootings_month_dummy$OCCUR_DATE, "%m"))
shootings_month_dummy$OCCUR_DATE <- NULL

# Merge the two datasets
shootings_month <- merge(
    shootings_month,
    shootings_month_dummy,
    by  = c("Borough", "Year", "Month"),
    all = TRUE # Produce NAN for missing values
)

# Fill missing values with 0
shootings_month$Shootings[is.na(shootings_month$Shootings)] <- 0
shootings_month$ShootingsDummy <- NULL
rm(shootings_month_dummy)

head(shootings_month)
```

The final dataset consists of $`r nrow(shootings_month)`$ rows and $`r ncol(shootings_month)`$ columns. Each column represents:

- $`r length(unique(shootings_month$Borough))`$ different boroughs in the `Borough` column.
- $`r length(unique(shootings_month$Month))`$ months in the `Month` column.
- $`r max(shootings_month$Year) - min(shootings_month$Year) + 1`$ years, from $`r min(shootings_month$Year)`$ to $`r max(shootings_month$Year)`$, in the `Year` column.
- The shootings count for each year, month, and borough in the `Shootings` column.

Here's the final summary of the dataset.

```{r shootings month summary}

summary(shootings_month)

```


## 3. Arrests data

Similarly to the shooting data, we load and preprocess the **arrests dataset** [`arrests.csv`](/data/arrests.csv) and create the columns suitable for temporal and spatial aggregation.

```{r load arrests}

arrests <- read.csv(PATHS$Arrests)

head(arrests)
```

In addition to the year, month, and borough, we also need to consider the specific **crime type** for which the arrest occurred. We focus on twelve crime types that are very frequent and potentially related to shootings. The crime types are as follows: `Assault`, `Burglary`, `DrugParaphernalia`, `Larceny`, `LicensingFirearm`, `MarijuanaPossession`, `Mischief`, `Murder`, `Robbery`, `SubstancePossession`, `Theft`, and `WeaponPossession`.


```{r arrests}

# Keep columns of interest
arrests <- arrests[, ARRESTS_COLS]

# Map categories to specified class
# and remove rows with other categories
for (crime in names(CRIME_TYPES)) {
    arrests$PD_DESC[arrests$PD_DESC %in% CRIME_TYPES[[crime]]] <- crime
}
arrests <- arrests[(arrests$PD_DESC %in% names(CRIME_TYPES)), ]

# Map boroughs to same name as shootings
arrests$ARREST_BORO <- BOROUGH_LETTER_MAP[arrests$ARREST_BORO]

# Cast to date 
arrests$ARREST_DATE  <- as.Date  (arrests$ARREST_DATE, format = "%m/%d/%Y")
arrests$Year         <- as.numeric(format(arrests$ARREST_DATE, "%Y"))
arrests$Month        <- as.numeric(format(arrests$ARREST_DATE, "%m"))
arrests$Borough      <- as.factor(arrests$ARREST_BORO)
arrests$Crime        <- as.factor(arrests$PD_DESC)
arrests$PD_DESC      <- NULL
arrests$ARREST_DATE  <- NULL
arrests$ARREST_BORO  <- NULL

head(arrests)
```

Below is a summary of the transformed dataset.

```{r}
summary(arrests)

```

We can now aggregate the arrests to obtain counts for each month, borough, and type of crime.

```{r arrests month aggregation}

# Create a new dataset counting how many arrests in each month
arrests_month <- aggregate(
    ARREST_KEY ~ Borough + Month + Year + Crime,
    data = arrests,
    FUN  = length
)
# Set count colname
names(arrests_month)[5] <- "Arrests"

# Merge with a dummy dataset to fill in missing months
dummy_arrests_month <- expand.grid(
    OCCUR_DATE   = seq(date.first, date.last, by = "month"),
    Borough      = levels(arrests$Borough),
    Crime        = levels(arrests$Crime),
    ArrestsDummy = 0
)

# Cast to the appropriate type
dummy_arrests_month$OCCUR_DATE <- as.Date          (dummy_arrests_month$OCCUR_DATE, format="%m/%d/%Y")
dummy_arrests_month$Year       <- as.numeric(format(dummy_arrests_month$OCCUR_DATE, "%Y"))
dummy_arrests_month$Month      <- as.numeric(format(dummy_arrests_month$OCCUR_DATE, "%m"))
dummy_arrests_month$OCCUR_DATE <- NULL

# Merge the two datasets
arrests_month <- merge(
    arrests_month,
    dummy_arrests_month,
    by  = c("Borough", "Year", "Month", "Crime"),
    all = TRUE # Produce NAN for missing values
)

# Fill missing values with 0
arrests_month$Arrests[is.na(arrests_month$Arrests)] <- 0
arrests_month$ArrestsDummy <- NULL

# Reset rows
rownames(arrests_month) <- NULL

head(arrests_month)

```

The final dataset consists of $`r nrow(arrests_month)`$ rows and $`r ncol(arrests_month)`$ columns. Each column represents:

- $`r length(unique(shootings_month$Borough))`$ different boroughs in column `Borough`.
- $`r length(unique(shootings_month$Month))`$ months in column `Month`.
- $`r max(shootings_month$Year) - min(shootings_month$Year) + 1`$ years, from $`r min(shootings_month$Year)`$ to $`r max(shootings_month$Year)`$, in column `Year`.
- $`r length(unique(shootings_month$Crimes))`$ different crime types in column `Crime`.
- The arrest count in column `Arrests` represents the number of arrests in the considered month, year, and borough for a specific crime type.

To make the dataset sutiable for merge with the shootings dataset, it is reshaped to create a column for each crime type. Moreover we add a column with the total number of arrests.

```{r arrests reshaping}

# Reorganaize the dataset using Crime as column
arrests_month <- reshape(
    arrests_month,
    idvar = c("Borough", "Year", "Month"),
    timevar = "Crime",
    direction = "wide"
)
names(arrests_month) <- gsub("Arrests.", "", names(arrests_month))

# Add a column with the total number of arrests
arrests_month$TotArrests <- rowSums(arrests_month[, 4:ncol(arrests_month)], na.rm = TRUE)

head(arrests_month)
```

The final summary of the arrests dataset follows.

```{r arrests month summary}

summary(arrests_month)

```

## 4. Merge datasets

We merge the two datasets to have a unique final view with the number of shootings and arrests for each borough, month and crime type. 

```{r merge datasets}

# Merge arrests and shootings
nyc_month_crimes <- merge(
    arrests_month,
    shootings_month, 
    by = c("Borough", "Year", "Month"), 
    all = TRUE
)

# Sort rows by Borough, Year and Month
nyc_month_crimes <- nyc_month_crimes[
    order(
        nyc_month_crimes$Borough, 
        nyc_month_crimes$Year,
        nyc_month_crimes$Month
    ), 
]

nyc_month_crimes$Borough <- BOROUGH_LOWERCASE_MAP[nyc_month_crimes$Borough]
nyc_month_crimes$Borough <- as.factor(nyc_month_crimes$Borough)

head(nyc_month_crimes)

```

The final dataset has $`r nrow(nyc_month_crimes)`$ rows and $`r ncol(nyc_month_crimes)`$ columns. Each row represent:

- $`r length(unique(shootings_month$Borough))`$ different boroughs in column `Borough`.
- $`r length(unique(shootings_month$Month))`$ months in column `Month`.
- $`r max(shootings_month$Year) - min(shootings_month$Year) + 1`$ years, from $`r min(shootings_month$Year)`$ to $`r max(shootings_month$Year)`$, in column `Year`.
- $`r length(unique(shootings_month$Crime))`$ different crime columns, including the synthetic one `TotArrests` describing the total number of arrests for that crime type.
- The `Shootings` feature represent the number of shootings in the considered area of the city and time period.


```{r save datasets}

summary(nyc_month_crimes)

# Save the datasets
write.csv(nyc_month_crimes, PATHS$NycMonthCrimes, row.names = FALSE)

```

We save the dataset for further analysis in the [`nyc_month_crimes.csv`](/data/nyc_month_crimes.csv) file.
