---
title: "NYC Shootings and Arrests - Data analysis"
author: "Sebastiano Quintavale - 878500"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: cerulean
        highlight: tango
---

## Introduction

```{r echo=FALSE, include=FALSE}

# 1. RMardown styling
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 8)

# 2. Project environment setup
source("src/init.R")

```


### Data preparation
```{r}

# Load data
crimes <- read.csv(PATHS$NycMonthCrimes)

```

The dataset has `r nrow(crimes)` rows and `r ncol(crimes)` columns.

```{r}

# Save the names of the crimes (excluding TotArrests)
crime_names <- colnames(crimes)[4:(length(colnames(crimes)) - 2)]

```

```{r}

# Add discrete version for Month and Year
crimes <- prepare_crimes(crimes_df=crimes, col_merge=FALSE)

summary(crimes)
head(crimes)

```

### Train and Test split

```{r}

# Split Train and Test
split <- train_test_split_by_year(
    df_=crimes, 
    year_test=YEAR_TEST
)

crimes      <- split$train
crimes_test <- split$test

rm(split)

```

## Visualizations

### Time series

```{r}

plot_multiple_ts(
    df_ts = ts_crimes_reshape(
        aggregate(
            . ~ Year + MonthName,
            data = crimes[, c("Year", "MonthName", crime_names, "Shootings")], # Remove Borough, Month, TotArrests
            FUN = sum
        )
    ),
    group  = "CrimeType",
    xlab   = "Year",
    ylab   = "Incidents",
    title  = "Criminality in NYC - 2006 to 2021 ",
    lwd    = 0.6,
    ticks  = 7,
    colors = PALETTE$crimes
)

```

```{r fig.width=16, fig.height=12}

BOROUGH_LEGEND <- extract_legend(

    plot_multiple_ts(
        df_ts = ts_borough_reshape(
            df_ = crimes[, c("Borough", "Year", "MonthName", "Shootings")],
            crime_type = "Shootings"
        ),
        group  = "Borough",
        xlab = "Year",
        ylab   = "Arrests",
        legend.direction = "horizontal",
        colors=unlist(PALETTE$boroughs)
    )

)

plot_in_grid <- function(plot_list, ncol = 1, title = "") {

    to_plot <- c(plot_list, list(BOROUGH_LEGEND))

    nplots <- length(plot_list)
    nrow <- ceiling(nplots / ncol)

    # Create the layout matrix
    layout_matrix <- matrix(seq(1, nrow * ncol), nrow = nrow, ncol = ncol, byrow = TRUE)
    layout_matrix <- rbind(layout_matrix, rep(max(layout_matrix, na.rm = TRUE) + 1, ncol))


    return(grid.arrange(
        grobs = to_plot,
        ncol = ncol,
        top = textGrob(
            title, 
            gp = gpar(fontsize = 15, fontface = "bold")),
            layout_matrix = layout_matrix,
            heights = c(rep(1, nrow), 0.2)  # Adjust the relative height for the legend row
    ))

}

ts_plots <- list()

for (crime_name in crime_names) {

    ts_plots[[crime_name]] <- plot_multiple_ts(
        df_ts = ts_borough_reshape(
            df_ = crimes[, c("Borough", "Year", "MonthName", crime_name)],
            crime_type = crime_name
        ),
        group  = "Borough",
        title  = paste(crime_name),
        lwd    = 0.6,
        colors = unlist(PALETTE$boroughs),
        ticks  = 3,
        legend = FALSE,
        title_size = 9
    )

}

plot_in_grid(
    plot_list = ts_plots,
    ncol = 5,
    title = "Crimes in NYC - 2006 to 2021"
)


```

```{r}

ts_plots <- list()

for (crime_name in c("TotArrests", "SqrtTotArrests")) {

    ts_plots[[crime_name]] <- plot_multiple_ts(
        df_ts = ts_borough_reshape(
            df_ = crimes[, c("Borough", "Year", "MonthName", crime_name)],
            crime_type = crime_name
        ),
        group  = "Borough",
        xlab   = "Year",
        ylab   = "Total Arrests",
        title  = paste(crime_name),
        lwd    = 0.6,
        colors = unlist(PALETTE$boroughs),
        ticks  = 5,
        legend = FALSE
    )

}

plot_in_grid(
    plot_list = ts_plots,
    ncol = 2,
    title = "Total Arrests in NYC - 2006 to 2021"
)

plot_multiple_ts(
    df_ts = ts_borough_reshape(
        df_ = crimes[, c("Borough", "Year", "MonthName", "Shootings")],
        crime_type = "Shootings"
    ),
    group  = "Borough",
    xlab   = "Year",
    ylab   = "Shootings Incidents",
    title  = "Shootings in NYC - 2006 to 2021",
    lwd    = 0.6,
    colors = unlist(PALETTE$boroughs),
    ticks  = 10,
)

```



### Principal Component Analysis

```{r fig.width = 8, fig.height = 6}

pca_analysis(
    df_       = crimes,
    features  = c(crime_names, "Shootings"),
    levels    = "Borough",
    title     = "PCA - Crimes in NYC"
)

pca_analysis(
    df_       = crimes[crimes$Borough != "StatenIsland", ],
    features  = c(crime_names, "Shootings"),
    levels    = "Borough",
    title     = "PCA - Crimes in NYC"
)

pca_analysis(
    df_       = crimes[!(crimes$Borough %in% c("Queens", "StatenIsland")), ],
    features  = c(crime_names, "Shootings"),
    levels    = "Borough",
    title     = "PCA - Crimes in NYC"
)
```

```{r}
crimes_v2 <- crimes[, -c(19)] %>%
    pivot_longer(
        cols = -(c(Borough, Year, Month, MonthName)),
        names_to = "CrimeType",
        values_to = "Count"
    ) %>%
    pivot_wider(
        names_from = Borough,
        values_from = Count
    )
crimes_v2$CrimeType <- as.factor(crimes_v2$CrimeType)

pca_analysis(
    df_       = crimes_v2,
    features  = levels(crimes$Borough),
    levels    = "CrimeType",
    title     = "PCA - Crimes in NYC"
)

rm(crimes_v2)

```

### Scatterplots
```{r}

BOROUGH_LEGEND <- extract_legend(

    levels_scatterplot(
        df_     = crimes,
        x       = "Year",
        y       = "Shootings",
        levels  = "Borough",
        legend.direction = "horizontal",
        colors = unlist(PALETTE$boroughs)
    )

)

scatterplots <- list()

# Plots
for (time in c("Year", "Month")) {
    scatterplots[[time]] <- levels_scatterplot(
        df_     = crimes,
        x       = time,
        y       = "Shootings",
        levels  = "Borough",
        title   = paste("Shootings by", time),
        jitter_ = TRUE,
        ticks   = ifelse(time=="Month", 12, 5),
        shape   = 19,
        size    = 1,
        colors  = unlist(PALETTE$boroughs),
        legend  = FALSE
    )
}

plot_in_grid(
    plot_list = scatterplots,
    ncol = 2,
    title = "Shootings in NYC over time"
)

```

```{r}

scatterplots <- list()

# Plots
for (crime in crime_names) {
    scatterplots[[crime]] <- levels_scatterplot(
        df_     = crimes,
        x       = crime,
        y       = "Shootings",
        levels  = "Borough",
        title   = paste(crime),
        title_size = 9,
        jitter_ = TRUE,
        ticks   = 3,
        shape   = 1,
        size    = .5,
        colors  = unlist(PALETTE$boroughs),
        legend  = FALSE
    )
}

plot_in_grid(
    plot_list = scatterplots,
    ncol = 5,
    title = "Crimes impact on shootings"
)

```

```{r}
scatterplots <- list()

# Plots
for (crime in c("TotArrests", "SqrtTotArrests")) {
    scatterplots[[crime]] <- levels_scatterplot(
        df_     = crimes,
        x       = crime,
        y       = "Shootings",
        levels  = "Borough",
        title   = paste(crime),
        title_size = 9,
        jitter_ = TRUE,
        ticks   = 3,
        shape   = 19,
        size    = 1.5,
        colors  = unlist(PALETTE$boroughs),
        legend  = FALSE
    )
}

plot_in_grid(
    plot_list = scatterplots,
    ncol = 2,
    title = "Crimes and shootings interaction"
)


```

Remarks:

- StatenInsland very few crimes and shootings
- In general Bronx and Brooklyn tends to behave similar, and Queens and Manhattan tends to behave similar.
- There's a decrease to 2020 and then raise
- There seasonal effect at different scales
- Most linear evidence: Assaults, Murder, Robbery
- Possible Bi-cluster for larceny
- Marijuana sell possession it's interesting but has initial outliers.
- Quadratic effect for SubstancePossession, WeaponPossesion and TotArrests

### Outliers

```{r}
# Extract outliers
crimes[
    crimes$TotArrests < 2000 &
        crimes$Shootings > 100,
]
```

Use tso to detect outliers in the time series ?
Treat as missing and use PCA to impute it ?


https://www.nytimes.com/2020/09/02/nyregion/nyc-shootings-murders.html

### Larceny clusters

```{r}

levels_scatterplot(
    df_     = crimes,
    x       = "Larceny",
    y       = "Shootings",
    xlab    = "Larceny",
    ylab    = "Shootings",
    levels  = "Borough",
    title   = "Larceny in NYC over years",
    title_size = 16,
    jitter_ = TRUE,
    ticks   = 10,
    shape   = 19,
    size    = 1.5,
    colors  = unlist(PALETTE$boroughs)
)

larceny_plots <- list()
for(level in levels(crimes$Borough)) {

    df_level <- crimes[crimes$Borough == level, ]
    df_level$Condition <- as.factor(ifelse(
        df_level$Year %in% 2014:2019, "A", "B"
    ))

    larceny_plots[[level]] <- levels_scatterplot(
        df_     = df_level,
        x       = "Larceny",
        y       = "Shootings",
        xlab    = "Larceny",
        ylab    = "Shootings",
        levels  = "Condition",
        title   = paste("Larceny - ", level),
        title_size = 12,
        jitter_ = TRUE,
        ticks   = 10,
        shape   = 19,
        size    = 1.5,
        colors  = c("blue", "orange"),
        legend = FALSE
    )
    
}

grid.arrange(
    grobs = larceny_plots,
    ncol = 2
)

```

### Marijuana outliers

```{r}

plot_in_grid(
    plot_list = list(
    marijuana = levels_scatterplot(
        df_     = crimes,
        x       = "MarijuanaSellPossession",
        y       = "Shootings",
        xlab    = "Arrests",
        ylab    = "MarijuanaSellPossession",
        levels  = "Borough",
        title   = "MarijuanaSellPossession in NYC over years",
        title_size = 12,
        jitter_ = TRUE,
        ticks   = 10,
        shape   = 19,
        size    = 1.5,
        legend  = FALSE,
        colors  = unlist(PALETTE$boroughs)
    ),
    substance = levels_scatterplot(
        df_     = crimes,
        x       = "SubstancePossession",
        y       = "Shootings",
        xlab    = "Arrests",
        ylab    = "SubstancePossession",
        levels  = "Borough",
        title   = "SubstancePossession in NYC over years",
        title_size = 12,
        jitter_ = TRUE,
        ticks   = 10,
        shape   = 19,
        size    = 1.5,
        legend  = FALSE,
        colors  = unlist(PALETTE$boroughs)
    )
),
    ncol = 2,
    title = ""
)

summary(as.factor(crimes[
    crimes$MarijuanaSellPossession == 0, 
]$Year))

```

```{r}
# Create a new column that is the sum of MarijuanaSellPOssesion and SubstancePossession
crimes$SubstanceAndMarijuanaPossession <- crimes$SubstancePossession + crimes$MarijuanaSellPossession

levels_scatterplot(
    df_     = crimes,
    x       = "SubstanceAndMarijuanaPossession",
    y       = "Shootings",
    xlab    = "Arrests",
    ylab    = "Shootings",
    levels  = "Borough",
    title   = "SubstanceAndMarijuanaPossession in NYC over years",
    title_size = 12,
    jitter_ = TRUE,
    ticks   = 10,
    shape   = 19,
    size    = 1.5,
    legend  = FALSE,
    colors  = unlist(PALETTE$boroughs)
)

crimes[
    crimes$SubstanceAndMarijuanaPossession <500 &
    crimes$Shootings > 50, 
    c("Borough", "Year", "SubstancePossession", "MarijuanaSellPossession", "Shootings")
]
```

### Crimes correlation

```{r}

full_model <- lm(
    Shootings ~ .,
    data = crimes[, c("Shootings", crime_names)]
)

summary(full_model)

plot(allEffects(full_model))

vif_diagnostic(full_model) 

```

```{r}
plot_correlation_matrix <- function(df) {

    # Compute the correlation matrix for numeric columns
    numeric_cols <- sapply(df, is.numeric)
    cor_matrix   <- cor(df[, numeric_cols], use = "complete.obs")

    # Visualize the correlation matrix using corrplot
    corrplot(
        cor_matrix, 
        method = "color",
        addCoef.col = "black",
        number.cex = 0.6, tl.cex = 0.6, cl.cex = 0.8
    )

}

plot_correlation_matrix(crimes[, c("Shootings", "TotArrests", crime_names)])
```

### Boroughs

```{r}
# Plot a boxplot for each borough
plot_boxplot <- function(df, x, y, title, colors=c()) {

    p = ggplot(df, aes(x = get(x), y = get(y))) +
        geom_boxplot(aes(fill = .data[[x]]), alpha = 0.7) +
        labs(title = title, x = x, y = y) +
        theme(legend.position = "none")

    if(length(colors)>0) {
        p <- p + scale_fill_manual(values = colors)
    }

    return(p)

}

plot_boxplot(
    crimes, "Borough", "Shootings", "Shootings by Borough",
    colors = unlist(PALETTE$boroughs)
)

crimes$Year <- as.factor(crimes$Year)
plot_boxplot(
    crimes, "Year", "Shootings", "Shootings by Year",
    colors = colorRampPalette(c('darkgreen', 'lightgreen'))(16)
)
crimes$Year <- as.numeric(crimes$Year)

plot_boxplot(
    crimes, "MonthName", "Shootings", "Shootings by Year",
    colors = colorRampPalette(c('#c44e00', '#eebf82'))(12)
)

```