---
title: "NYC Shootings and Arrests - Data preprocessing"
author: "Sebastiano Quintavale - 878500"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    html_document:
        toc: true
    pdf_document:
        toc: true
---

## 1. Introduction

This document describes the data preprocessing phase of the project. The goal is to clean and transform the raw data into a format that can be used for the analysis. The data used in this project are the following:

- **NYPD Shooting Incident Data**: contains information about shooting incidents in New York City in the period 2006-2023. The dataset is available at NYC Data portal at the follwing [resource](https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8).
- **NYPD Arrest Data**: contains information about arrests in New York City in the period 2006-2023. The dataset is available at NYC Data portal at the following [resource](https://data.cityofnewyork.us/Public-Safety/NYPD-Arrest-Data-Historic-/8h9b-rp9u).

The preprocissing has the goal of build a monthly dataset specifying the number of **shootings** and arrests for each borough and **crime type**. 

```{r setup, echo=FALSE}

# Don't display code in the final document
knitr::opts_chunk$set(echo=FALSE)

# Set English language
Sys.setlocale("LC_TIME", "en_US.UTF-8")

# Clear the environment
rm(list = ls())

# Load libraries
library("chron") # manipulate timedays

# Load R script sources
source("src/settings.R")
```

```{r names}

SHOOTINGS_COLS <- c("INCIDENT_KEY", "OCCUR_DATE",  "BORO")
ARRESTS_COLS   <- c("ARREST_KEY",   "ARREST_DATE", "ARREST_BORO", "PD_DESC")

# Name mappings

BOROUGH_LETTER_MAP <- c(
    "S" = "STATEN ISLAND",
    "K" = "BROOKLYN",
    "M" = "MANHATTAN",
    "B" = "BRONX",
    "Q" = "QUEENS"
)

BOROUGH_LOWERCASE_MAP <- c(
    "STATEN ISLAND" = "StatenIsland",
    "BROOKLYN"      = "Brooklyn",
    "MANHATTAN"     = "Manhattan",
    "BRONX"         = "Bronx",
    "QUEENS"        = "Queens",
    "NEW YORK"      = "NewYork"
)

CRIME_TYPES <- list(

    # Aggressione
    Assaults = c(
        "ASSAULT 3",                 "ASSAULT 2,1,UNCLASSIFIED", 
        "ASSAULT 2,1,PEACE OFFICER", "ASSAULT POLICE/PEACE OFFICER"
    ),

    # Possesso e vendita di marijuana
    MarijuanaSellPossession = c(  
        "MARIJUANA, POSSESSION 4 & 5", "MARIJUANA, SALE 4 & 5",    "MARIJUANA, POSSESSION 1, 2 & 3",
        "MARIJUANA, POSSESSION",       "MARIJUANA, SALE 1, 2 & 3", "UNLAWFUL SALE SYNTHETIC MARIJUANA"
    ),

    # Furto
    Theft = c( 
        "THEFT OF SERVICES, UNCLASSIFIED",     "THEFT OF SERVICES, UNCLASSIFIE",  "THEFT,RELATED OFFENSES,UNCLASSIFIED",
        "THEFT OF SERVICES- CABLE TV SERVICE", "THEFT,RELATED OFFENSES,UNCLASS"
    ),

    # Arma da fuoco
    Firearm = c( 
        'CRIMINAL DISPOSAL FIREARM 1 & 2', "CRIMINAL DISPOSAL FIREARM 1 &", "CRIMINAL DISPOSAL FIREARM 1",
        "LICENSING FIREARMS",              "FIREARMS LICENSING LAWS"
    ),

    # Possesso/vendita di sostanza controllata
    SubstancePossession = c(  
        "CONTROLLED SUBSTANCE, POSSESSION 7", "CONTROLLED SUBSTANCE,INTENT TO SELL 3",  "CONTROLLED SUBSTANCE,SALE 3",
        "CONTROLLED SUBSTANCE, POSSESSI",     "CONTROLLED SUBSTANCE, INTENT TO SELL 5", "CONTROLLED SUBSTANCE,INTENT TO",
        "CONTROLLED SUBSTANCE, POSSESSION 5", "CONTROLLED SUBSTANCE,POSSESS. 1",        "CONTROLLED SUBSTANCE, POSSESSION 4",
        "CONTROLLED SUBSTANCE,POSSESS. 2",    "CONTROLLED SUBSTANCE, SALE 5",           "CONTROLLED SUBSTANCE,POSSESS.",
        "CONTROLLED SUBSTANCE, INTENT T",     "CONTROLLED SUBSTANCE,SALE 1",            "CONTROLLED SUBSTANCE,POSSESS. 3",
        "CONTROLLED SUBSTANCE,SALE 2",        "CONTROLLED SUBSTANCE, SALE 4",           "CONTROLLED SUBSTANCE,POSSESS. OF PROCURSERS"
    ),

    # Rapina
    Robbery = c( 
        "ROBBERY,UNCLASSIFIED,OPEN AREAS",           "ROBBERY,OPEN AREA UNCLASSIFIED", "ROBBERY,UNCLASSIFIED,OPEN AREA",
        "ROBBERY,CARJACKING OF MV OTHER THAN TRUCK", "ROBBERY,CAR JACKING",            "ROBBERY,GAS STATION"
    ),

    # Possesso di armi
    WeaponsPossession = c( 
        "WEAPONS, POSSESSION, ETC",   "WEAPONS POSSESSION 3",                    "WEAPONS POSSESSION 1 & 2",
        "WEAPONS,MFR,TRANSPORT,ETC.", "WEAPONS,PROHIBITED USE",                  "WEAPONS,DISPOSITION OF",
        "WEAPONS DISPOSITION OF",     "WEAPONS,PROHIBITED USE IMITATION PISTOL", "WEAPONS,PROHIBITED USE IMITATI"
    ),

    # Legami per droga
    DrugParaphernalia = c(  
        "DRUG PARAPHERNALIA,   POSSESSES OR SELLS 2", "DRUG PARAPHERNALIA,   POSSESSE",
        "DRUG PARAPHERNALIA,   POSSESSES OR SELLS 1"
    ),

    # Furto
    Larceny= c(
        "LARCENY,PETIT FROM OPEN AREAS,UNCLASSIFIED", "LARCENY,PETIT FROM OPEN AREAS,",           "LARCENY,GRAND FROM OPEN AREAS,UNCLASSIFIED",
        "LARCENY,GRAND FROM OPEN AREAS, UNATTENDED",  "LARCENY,GRAND OF AUTO",                    "LARCENY,GRAND FROM PERSON,UNCLASSIFIED",
        "LARCENY,GRAND FROM PERSON,UNCL",             "LARCENY,GRAND FROM OPEN AREAS,",           "LARCENY,GRAND BY EXTORTION",
        "AGGRAVATED GRAND LARCENY OF ATM",            "LARCENY,GRAND BY CREDIT CARD USE",         "LARCENY,PETIT BY ACQUIRING LOS",
        "LARCENY,GRAND BY ACQUIRING LOS",             "LARCENY,GRAND FROM BUILDING,UNCLASSIFIED", "LARCENY,GRAND BY THEFT OF CREDIT CARD",
        "LARCENY,GRAND FROM BUILDING (NON-RESIDENCE) UNATTENDED"
    ),

    # Frode
    Mischief = c( 
        "MISCHIEF,CRIMINAL     UNCLASSIFIED 4TH DEG", "MISCHIEF,CRIMINAL,    UNCL 2ND",             "MISCHIEF,CRIMINAL,    UNCL 2ND DEG 3RD DEG",
        "CRIMINAL MISCHIEF,UNCLASSIFIED 4",           "CRIMINAL MISCHIEF 4TH, GRAFFITI",            "CRIMINAL MISCHIEF 4TH, GRAFFIT",
        "MISCHIEF, CRIMINAL 4, OF MOTOR VEHICLE",     "MISCHIEF, CRIMINAL 3 & 2, OF MOTOR VEHICLE", "MISCHIEF 1,CRIMINAL,EXPLOSIVE",
        "MISCHIEF, CRIMINAL 4, BY FIRE",              "MISCHIEF, CRIMINAL 3 & 2, OF M",             "MISCHIEF, CRIMINAL 4, OF MOTOR",
        "MISCHIEF,CRIMINAL     UNCLASSI"
    ),

    # Trasgressione
    Trepass = c(  
        "TRESPASS 3, CRIMINAL", "TRESPASS 2, CRIMINAL", "TRESPASS 4,CRIMINAL",
        "TRESPASS 1,CRIMINAL",  "TRESPASS 4,CRIMINAL SUB 2"
    ),

    # Furto con scasso
    Burglary = c(  
        "BURGLARY,UNCLASSIFIED,UNKNOWN TIME", "BURGLARY,UNCLASSIFIED,UNKNOWN",
        "BURGLARY,RESIDENCE,NIGHT",           "BURGLARY,COMMERCIAL,DAY"
    ),

    # Corruzione
    Bribery = c(  
        "BRIBERY,PUBLIC ADMINISTRATION", "BRIBERY,COMMERCIAL", "BRIBERY,FRAUD"
    ),

    # Omicidio
    Murder = c("MURDER,UNCLASSIFIED"),

    # Falsificazione
    Forgery = c(
        "FORGERY,ETC.,UNCLASSIFIED-FELONY", "FORGERY,ETC.-MISD.",                           "FORGERY,ETC.,UNCLASSIFIED-FELO", "FORGERY,M.V. REGISTRATION", 
        "FORGERY-ILLEGAL POSSESSION,VEH",   "FORGERY-ILLEGAL POSSESSION,VEHICLE IDENT. NU", "FORGERY,PRESCRIPTION"
    )
)


```

## 2. Shootings data

We load and preprocess the **shootings dataset** [`shootings.csv`](/data/shooting.csv) and we crate the columns suitable for aggregation purposes.

```{r load shootings}

shootings <- read.csv(FILE_PATHS$Shootings)

```

```{r shootings type casting}

# Keep columns of interest
shootings <- shootings[, SHOOTINGS_COLS]

# Cast to date and time type
shootings$OCCUR_DATE <- as.Date(shootings$OCCUR_DATE, format = "%m/%d/%Y")

# Cast Borough to factor
shootings$Borough       <- as.factor(shootings$BORO)
shootings$BORO          <- NULL

# Extract time features
shootings$Year    <- as.numeric(format(shootings$OCCUR_DATE, "%Y"))
shootings$Month   <- as.numeric(format(shootings$OCCUR_DATE, "%m"))

# Save the first and last date
date.first <- min(shootings$OCCUR_DATE)
date.last  <- max(shootings$OCCUR_DATE)

shootings$OCCUR_DATE <- NULL

head(shootings)
summary(shootings)
```

We aggregate the shootings data by year, month and borough to have a count of shootings for each month and borough.

```{r shootings month aggregation}

# Create a new dataset counting how many arrests in each month
shootings_month <- aggregate(
    INCIDENT_KEY ~ Borough + Month + Year,
    data = shootings,
    FUN  = length
)

# Change count colname
names(shootings_month)[4] <- "Shootings"

# Create a dummy dataset to fill missing months with no shooting
shootings_month_dummy <- expand.grid(
    OCCUR_DATE     = seq(date.first, date.last, by = "month"),
    Borough        = levels(shootings$Borough),
    ShootingsDummy = 0
)

# Cast to the appropriate type
shootings_month_dummy$OCCUR_DATE <- as.Date(shootings_month_dummy$OCCUR_DATE, format="%m/%d/%Y")
shootings_month_dummy$Year       <- as.numeric(format(shootings_month_dummy$OCCUR_DATE, "%Y"))
shootings_month_dummy$Month      <- as.numeric(format(shootings_month_dummy$OCCUR_DATE, "%m"))
shootings_month_dummy$OCCUR_DATE <- NULL

# Merge the two datasets
shootings_month <- merge(
    shootings_month,
    shootings_month_dummy,
    by  = c("Borough", "Year", "Month"),
    all = TRUE # Produce NAN for missing values
)

# Fill missing values with 0
shootings_month$Shootings[is.na(shootings_month$Shootings)] <- 0
shootings_month$ShootingsDummy <- NULL
rm(shootings_month_dummy)

head(shootings_month)
```

The final dataset consists of $1080$ rows and $4$ columns. Each row represent:

- 5 different boroughs in column `Borough`.
- 12 months in column `Month`.
- 18 years, from 2006 to 2023, in column `Year`

The feature `Shootings` represent the number of shootings in the considered area and period.

Here's the final summary of the dataset.

```{r shootings month summary}

summary(shootings_month)

```


# 3. Arrests data

As we did for the shooting data, we load and preprocess the **arrests dataset** [`arrests.csv`](/data/arrests.csv) and we crate the columns suitable for aggregation purposes. In addition to the year, month and borough, we also need to take in consideration the specific **crime type** the arrest occured.
We only keep the type of crimes most frequent fifteen and possibily related with shootings. The crime types are the following : Assuault, Marijuana possesion and sell, Theft, Firearm, SubstancePossession, Robbery, Weapons possesion, Drug paraphernalia, Larceny, Mischief, Trepass, Burglary, Bribery, Murder and Forgery.


```{r load arrests}

arrests <- read.csv(FILE_PATHS$Arrests)

```

```{r arrests}

# Keep columns of interest
arrests <- arrests[, ARRESTS_COLS]

# Map categories to specified class
# and remove rows with other categories
for (crime in names(CRIME_TYPES)) {
    arrests$PD_DESC[arrests$PD_DESC %in% CRIME_TYPES[[crime]]] <- crime
}
arrests <- arrests[(arrests$PD_DESC %in% names(CRIME_TYPES)), ]

# Map boroughs to same name as shootings
arrests$ARREST_BORO <- BOROUGH_LETTER_MAP[arrests$ARREST_BORO]

# Cast to date 
arrests$ARREST_DATE  <- as.Date  (arrests$ARREST_DATE, format = "%m/%d/%Y")
arrests$Year         <- as.numeric(format(arrests$ARREST_DATE, "%Y"))
arrests$Month        <- as.numeric(format(arrests$ARREST_DATE, "%m"))
arrests$Borough      <- as.factor(arrests$ARREST_BORO)
arrests$Crime        <- as.factor(arrests$PD_DESC)
arrests$PD_DESC      <- NULL
arrests$ARREST_DATE  <- NULL
arrests$ARREST_BORO  <- NULL

head(arrests)

```

We aggregate the arrests to have a count of arrests for each month, borough and crime.

```{r arrests month aggregation}

# Create a new dataset counting how many arrests in each month
arrests_month <- aggregate(
    ARREST_KEY ~ Borough + Month + Year + Crime,
    data = arrests,
    FUN  = length
)
# Set count colname
names(arrests_month)[5] <- "Arrests"

# Merge with a dummy dataset to fill in missing months
dummy_arrests_month <- expand.grid(
    OCCUR_DATE   = seq(date.first, date.last, by = "month"),
    Borough      = levels(arrests$Borough),
    Crime        = levels(arrests$Crime),
    ArrestsDummy = 0
)

# Cast to the appropriate type
dummy_arrests_month$OCCUR_DATE <- as.Date          (dummy_arrests_month$OCCUR_DATE, format="%m/%d/%Y")
dummy_arrests_month$Year       <- as.numeric(format(dummy_arrests_month$OCCUR_DATE, "%Y"))
dummy_arrests_month$Month      <- as.numeric(format(dummy_arrests_month$OCCUR_DATE, "%m"))
dummy_arrests_month$OCCUR_DATE <- NULL

# Merge the two datasets
arrests_month <- merge(
    arrests_month,
    dummy_arrests_month,
    by  = c("Borough", "Year", "Month", "Crime"),
    all = TRUE # Produce NAN for missing values
)

# Fill missing values with 0
arrests_month$Arrests[is.na(arrests_month$Arrests)] <- 0
arrests_month$ArrestsDummy <- NULL

head(arrests_month)

```

The final dataset consists of $16200$ rows and $5$ columns. Each row represent:

- 5 different boroughs in column `Borough`.
- 12 months in column `Month`.
- 18 years, from 2006 to 2023, in column `Year`
- 15 different crime types in column `Crime`

The `Arrests` feature represent the number of arrests in the considered area and period.

However we prefer to rehsape the dataset to use the crime type as column in order to make the dataset suitable for the merge with shooting one.


```{r arrests reshaping}

# Reorganaize the dataset using Crime as column
arrests_month <- reshape(
    arrests_month,
    idvar = c("Borough", "Year", "Month"),
    timevar = "Crime",
    direction = "wide"
)
names(arrests_month) <- gsub("Arrests.", "", names(arrests_month))

# Add a column with the total number of arrests
arrests_month$TotArrests <- rowSums(arrests_month[, 4:ncol(arrests_month)], na.rm = TRUE)

head(arrests_month)

```

The final summary of the dataset is the following.

```{r arrests month summary}

summary(arrests_month)

```

## 4. Merge datasets

We merge the two datasets to have a unique dataset with the number of shootings and arrests for each borough, month and crime type. We also create an artificial borough **NewYork** accounting for the total number of shootings and crimes in the city.

```{r merge datasets}

# Merge arrests and shootings
nyc_month_crimes <- merge(
    arrests_month,
    shootings_month, 
    by = c("Borough", "Year", "Month"), 
    all = TRUE
)

# Sort rows by Borough, Year and Month
nyc_month_crimes <- nyc_month_crimes[
    order(
        nyc_month_crimes$Borough, 
        nyc_month_crimes$Year,
        nyc_month_crimes$Month
    ), 
]

nyc_month_crimes$Borough <- BOROUGH_LOWERCASE_MAP[nyc_month_crimes$Borough]
nyc_month_crimes$Borough <- as.factor(nyc_month_crimes$Borough)

head(nyc_month_crimes)

```

The final dataset has $1080$ rows and $20$ columns. Each row represent:

- 5 different boroughs in column `Borough`.
- 12 months in column `Month`.
- 18 years, from 2006 to 2023, in column `Year`.
- 15 different crime types, plus their total in column `TotArrests`.
- The `Shootings` feature represent the number of shootings in the considered area and period.

```{r save datasets}

summary(nyc_month_crimes)

# Save the datasets
write.csv(nyc_month_crimes, FILE_PATHS$NycMonthCrimes, row.names = FALSE)

```

We save the dataset for the analysis in the [`nyc_month_crimes.csv`](/data/nyc_month_crimes.csv) file.
